**Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces**

### 2020-03-24 submission/manuscript.Rmd
References checked.

\vspace{50mm}
Marino Korlević^1$*$^, Marsej Markovski^1^, Zihao Zhao^2^, Gerhard J. Herndl^2,3^, Mirjana Najdek^1^

\vspace{40mm}

^$*$^To whom correspondence should be addressed: marino.korlevic@irb.hr

@@

@@

## Abstract
Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited by diverse microbial communities. Most studies focusing on macrophyte epiphytic communities did not take into account temporal changes or applied low sampling frequency approaches. Illumina sequencing of the V4 16S rRNA region was performed to determine the seasonal dynamics of epiphytic communities sampled from the surfaces of the seagrass *Cymodocea nodosa* and invasive macroalga *Caulerpa cylindracea*. Leaves and thalli were sampled in a meadow of *Cymodocea nodosa* invaded by the invasive *Caulerpa cylindracea* and in a monospecific settlement of *Caulerpa cylindracea* located in the proximity of the meadow at monthly intervals. For comparison the ambient prokaryotic plankton community was also characterized. Sequencing results at the OTU level showed a clear differentiation between ambient water and epiphytic communities and a host-specific community assemblage. In addition, successional changes were observed that could be connected to the macrophyte growth cycle. Taxonomic analysis showed similar high rank groups in the ambient water and epiphytic communities, with the exception of *Desulfobacterota* that were found only on *Caulerpa cylindracea*. Only *Cyanobacteria* showed seasonal change, while other high rank taxa were present throughout the year. In every analyzed high rank taxa, phylogenetic groups present throughout the year comprised most of the sequences and could be identified together with low proportion taxa showing seasonal patterns connected to the macrophyte growth cycle. Taken together, epiphytic microbial communities of the seagrass *Cymodocea nodosa* and the macroalgae *Caulerpa cylindracea* appear to be host-specific and contain taxa that undergo successional changes.

@@

All these ecological roles are carried out by a taxonomically diverse community of microorganisms. At higher taxonomic ranks a core set of macrophyte epiphytic taxa was described consisting mainly of *Alphaproteobacteria*, *Gammaproteobacteria*, *Desulfobacterota*, *Bacteroidota*, *Cyanobacteria*, *Actinobacteriota*, *Firmicutes*, *Planctomycetota*, *Chloroflexi* and *Verrucomicrobiota* [@Lachnit2011; @Crump2008; @Tujula2010]. In contrast, at lower taxonomic ranks host specific microbial communities were described [@Roth-Schulze2016; @Lachnit2011]. Recently, it was shown that even different morphological niches within the same alga had a higher influence on bacterial community variation than biogeography or environmental factors [@Morrissey2019]. While there is high community variation between host species it was observed that the majority of metagenome determined functions were conserved both between host species and individuals [@Burke2011a; @Roth-Schulze2016]. This discrepancy between taxonomic and functional composition could be explained by the lottery hypothesis. It postulates that an initial random colonization step is performed from a set of functionally equivalent taxonomic groups resulting in taxonomically different epiphytic communities sharing a core set of functional genes [@Burke2011a; @Roth-Schulze2016]. In addition, some of the variation in the observed data could be attributed to different techniques used in various studies, such as different protocols for epiphytic cell detachment and/or DNA isolation, as no standard protocol to study epiphytic communities was established [@Ugarelli2019; @Korlevic_submitted, submitted].

@@

\newpage
## Discussion
Surfaces of marine macrophytes harbor biofilms consisting of diverse microbial taxa [@Egan2013; @Tarquinio2019]. No standard protocol has been developed to study these macophyte associated microbes [@Ugarelli2019]. Different procedures for removal of microbial cells from host surfaces were described, such as host tissue shaking [@Noges2010], scraping [@Uku2007] and ultrasonication [@Cai2014]. All these methods showed different removal efficiencies but none was enabling a complete removal of attached microbial cells. In the present study, we applied an earlier developed removal protocol [@Korlevic_submitted, submitted], based on a previous idea of direct cellular lysis [@Burke2009], to ensure an almost complete cell detachment. The application of a direct lysis procedure coupled with a high frequency sampling protocol and Illumina high resolution amplicon sequencing has enabled us to make a detailed description of bacterial and archaeal communities associated with the surfaces of two marine macrophytes, *C. nodosa* and *C. cylindracea*.

@@

\newpage
## Experimental Procedures
### Sampling
Sampling was performed in the Bay of Funtana, northern Adriatic Sea (\ang{45;10;39} N, \ang{13;35;42} E). Leaves and thalli were sampled in a meadow of *Cymodocea nodosa* invaded by the invasive *Caulerpa cylindracea* (mixed settlement) and in a monospecific settlement of *Caulerpa cylindracea* located in the proximity of the meadow at approximately monthly intervals from December 2017 to October 2018 (\autoref{supp-nseq_notus}). Leaves and thalli were collected by diving and transported to the laboratory in containers placed on ice and filled with site seawater. Upon arrival to the laboratory, *C. nodosa* leaves were cut into sections of 1 -- 2 \si{\cm}, while *C. cylindracea* thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three times with sterile artificial seawater (ASW) to remove loosely attached microbial cells. Ambient seawater was collected in 10 \si{\l} containers by diving and transported to the laboratory where the whole container volume was filtered through a 20 \si{\um} net. The filtrate was further sequentially filtered through 3 \si{\um} and 0.2 \si{\um} polycarbonate membrane filters (Whatman, United Kingdom) using a peristaltic pump. Filters were briefly dried at room temperature and stored at \num{-80} \si{\degreeCelsius}. Seawater samples were also collected approximately monthly from July 2017 to October 2018.

### DNA Isolation
DNA from surfaces of *C. nodosa* and *C. cylindracea* was isolated using a previously modified and adapted protocol that allows for a selective epiphytic DNA isolation [@Korlevic_submitted, submitted; @Massana1997]. Briefly, leaves and thalli are incubated in a lysis buffer and treated with lysozyme and proteinase K. Following the incubations, the mixture containing lyzed epiphytic cells is separated from leaves and thalli and extracted using a phenol-chloroform procedure. Finally, the extracted DNA is precipitated using isopropanol. DNA from seawater picoplankton was isolated from 0.2 \si{\um} polycarbonate filters according to @Massana1997 with a slight modification. Following the phenol-chloroform extraction steps 1/10 of chilled 3 \si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1 volume of chilled isopropanol, incubating the mixtures overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 20,000 × g and 4 \si{\degreeCelsius} for 21 \si{\minute}. The pellet was washed twice with 500 \si{\ul} of chilled 70 \si{\percent} ethanol and centrifuged after each washing step at 20,000 × g and 4 \si{\degreeCelsius} for 5 \si{\minute}. Dried pellets were resuspended in 50 -- 100 \si{\ul} of deionized water.

### Illumina 16S rRNA Sequencing
Illumina MiSeq sequencing of the V4 16S rRNA region was performed as described previously [@Korlevic_submitted, submitted]. The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5' end a tagged sequence. Purified PCR products were sent for Illumina MiSeq sequencing at IMGM Laboratories, Martinsried, Germany. Before sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. Beside samples, a positive and negative control for each sequencing batch was sequenced. Negative control was comprised of PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

### Sequence Analysis
```{r error}
# Calculation of the error rate based on the ATCC MSA-1002 mock community
error <- read_tsv("data/mothur/raw.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.error.summary", na="") %>%
  filter(numparents==1) %>%
  mutate(mismatches_weight=weight*mismatches) %>%
  mutate(total_weight=weight*total)
error <- sum(error$mismatches_weight)/sum(error$total_weight)*100

@@

## Figure Captions
**\autoref{matrix}.** \nameref{matrix}

>>>>

\vspace{20mm}
Marino Korlević^1$*$^, Marsej Markovski^1^, Zihao Zhao^2^, Gerhard J. Herndl^2,3^, Mirjana Najdek^1^

@@

^$*$^To whom correspondence should be addressed:

Marino Korlević

G. Paliaga 5, 52210 Rovinj, Croatia

Tel.: +385 52 804 768

Fax: +385 52 804 780

e-mail: marino.korlevic@irb.hr

Running title: Seasonal dynamics of epiphytic communities

@@

## Summary
Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited by diverse microbial communities. Most studies focusing on macrophyte epiphytic communities did not take into account temporal changes or applied low sampling frequency approaches. Illumina sequencing of the V4 16S rRNA region was performed to determine the seasonal dynamics of epiphytic communities sampled from the surfaces of the seagrass *Cymodocea nodosa* and invasive macroalga *Caulerpa cylindracea*. Leaves and thalli were sampled in a meadow of *Cymodocea nodosa* invaded by the invasive *Caulerpa cylindracea* and in a monospecific settlement of *Caulerpa cylindracea* located in the proximity of the meadow at monthly intervals. For comparison the ambient prokaryotic plankton community was also characterized. Sequencing results at the OTU level showed a clear differentiation between ambient water and epiphytic communities and a host-specific community assemblage. In addition, successional changes were observed that could be connected to the macrophyte growth cycle. Taxonomic analysis showed similar high rank groups in the ambient water and epiphytic communities, with the exception of *Desulfobacterota* that were found only on *Caulerpa cylindracea*. Only *Cyanobacteria* showed seasonal change, while other high rank taxa were present throughout the year. In every analyzed high rank taxa, phylogenetic groups present throughout the year comprised most of the sequences and could be identified together with low proportion taxa showing seasonal patterns connected to the macrophyte growth cycle. Taken together, epiphytic microbial communities of the seagrass *Cymodocea nodosa* and the macroalgae *Caulerpa cylindracea* appear to be host-specific and contain taxa that undergo successional changes.

@@

All these ecological roles are carried out by a taxonomically diverse community of microorganisms. At higher taxonomic ranks a core set of macrophyte epiphytic taxa was described consisting mainly of *Alphaproteobacteria*, *Gammaproteobacteria*, *Desulfobacterota*, *Bacteroidota*, *Cyanobacteria*, *Actinobacteriota*, *Firmicutes*, *Planctomycetota*, *Chloroflexi* and *Verrucomicrobiota* [@Lachnit2011; @Crump2008; @Tujula2010]. In contrast, at lower taxonomic ranks host specific microbial communities were described [@Roth-Schulze2016; @Lachnit2011]. Recently, it was shown that even different morphological niches within the same alga had a higher influence on bacterial community variation than biogeography or environmental factors [@Morrissey2019]. While there is high community variation between host species it was observed that the majority of metagenome determined functions were conserved both between host species and individuals [@Burke2011a; @Roth-Schulze2016]. This discrepancy between taxonomic and functional composition could be explained by the lottery hypothesis. It postulates that an initial random colonization step is performed from a set of functionally equivalent taxonomic groups resulting in taxonomically different epiphytic communities sharing a core set of functional genes [@Burke2011a; @Roth-Schulze2016]. In addition, some of the variation in the observed data could be attributed to different techniques used in various studies, such as different protocols for epiphytic cell detachment and/or DNA isolation, as no standard protocol to study epiphytic communities was established [Korlević *et al.*, unpublished; @Ugarelli2019].

@@

\newpage
## Discussion
Surfaces of marine macrophytes harbor biofilms consisting of diverse microbial taxa [@Egan2013; @Tarquinio2019]. No standard protocol has been developed to study these macophyte associated microbes [@Ugarelli2019]. Different procedures for removal of microbial cells from host surfaces were described, such as host tissue shaking [@Noges2010], scraping [@Uku2007] and ultrasonication [@Cai2014]. All these methods showed different removal efficiencies but none was enabling a complete removal of attached microbial cells. In the present study, we applied an earlier developed removal protocol (Korlević *et al.*, unpublished), based on a previous idea of direct cellular lysis [@Burke2009], to ensure an almost complete cell detachment. The application of a direct lysis procedure coupled with a high frequency sampling protocol and Illumina high resolution amplicon sequencing has enabled us to make a detailed description of bacterial and archaeal communities associated with the surfaces of two marine macrophytes, *C. nodosa* and *C. cylindracea*.

@@

\newpage
## Experimental procedures
### Sampling
Sampling was performed in the Bay of Funtana, northern Adriatic Sea (\ang{45;10;39} N, \ang{13;35;42} E). Leaves and thalli were sampled in a meadow of *Cymodocea nodosa* invaded by the invasive *Caulerpa cylindracea* (mixed settlement) and in a monospecific settlement of *Caulerpa cylindracea* located in the proximity of the meadow at approximately monthly intervals from December 2017 to October 2018 (\autoref{supp-nseq_notus}). Leaves and thalli were collected by diving and transported to the laboratory in containers placed on ice and filled with site seawater. Upon arrival to the laboratory, *C. nodosa* leaves were cut into sections of 1 -- 2 \si{\cm}, while *C. cylindracea* thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three times with sterile artificial seawater (ASW) to remove loosely attached microbial cells. Ambient seawater was collected in 10 \si{\l} containers by diving and transported to the laboratory where the whole container volume was filtered through a 20 \si{\um} net. The filtrate was further sequentially filtered through 3 \si{\um} and 0.2 \si{\um} polycarbonate membrane filters (Whatman, United Kingdom) using a peristaltic pump. Filters were briefly dried at room temperature and stored at \num{-80} \si{\degreeCelsius}. Seawater samples were also collected approximately monthly from July 2017 to October 2018.

### DNA isolation
DNA from surfaces of *C. nodosa* and *C. cylindracea* was isolated using a previously modified and adapted protocol that allows for a selective epiphytic DNA isolation [Korlević *et al.*, unpublished; @Massana1997]. Briefly, leaves and thalli are incubated in a lysis buffer and treated with lysozyme and proteinase K. Following the incubations, the mixture containing lyzed epiphytic cells is separated from leaves and thalli and extracted using a phenol-chloroform procedure. Finally, the extracted DNA is precipitated using isopropanol. DNA from seawater picoplankton was isolated from 0.2 \si{\um} polycarbonate filters according to @Massana1997 with a slight modification. Following the phenol-chloroform extraction steps 1/10 of chilled 3 \si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1 volume of chilled isopropanol, incubating the mixtures overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 20,000 × g and 4 \si{\degreeCelsius} for 21 \si{\minute}. The pellet was washed twice with 500 \si{\ul} of chilled 70 \si{\percent} ethanol and centrifuged after each washing step at 20,000 × g and 4 \si{\degreeCelsius} for 5 \si{\minute}. Dried pellets were resuspended in 50 -- 100 \si{\ul} of deionized water.

### Illumina 16S rRNA sequencing
Illumina MiSeq sequencing of the V4 16S rRNA region was performed as described previously (Korlević *et al.*, unpublished). The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5' end a tagged sequence. Purified PCR products were sent for Illumina MiSeq sequencing at IMGM Laboratories, Martinsried, Germany. Before sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. Beside samples, a positive and negative control for each sequencing batch was sequenced. Negative control was comprised of PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

### Sequence analysis
```{r error}
# Calculation of the error rate based on the ATCC MSA-1002 mock community
error <- read_tsv("data/mothur/raw.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.error.summary", na="") %>%
  filter(numparents==1) %>%
  mutate(mismatches_weight=weight*mismatches) %>%
  mutate(total_weight=weight*total)
error <- sum(error$mismatches_weight)/sum(error$total_weight)*100

@@

## Figure legends
**\autoref{matrix}.** \nameref{matrix}
